Scheduling Musical Events in Max/MSP with Scheme For Max

### 2021-08-09 paper-1/paper-1.tex
finished revisions

In both the fields of live-coding and algorithmic music, the ability for
the performer and composer to schedule events in the future with high temporal
accuracy is of major benefit. While this is possible in the Max visual
patching language, it is cumbersome and thus the implementation is not
optimal for live-coding or algorithmic music where the speed with which
the user can accomplish this is an important consideration (i.e., fast
enough not to bore an audience or to be frustrating while composing).

@@

Activity in an object (represented visually by a single visual box in the patch) is triggered
by sending the object a Max message, most commonly from an object connected to it
through a patch cord running to an ``inlet" of the object. 
In order for the receiving object to do anything, it must 
be sent a message with a message selector for which it has a bound method,
a situation referred to in the nomenclature as ``responding to the message". Most objects
have a principal activity that typically ends with outputting the result of their calculation, and this is
often triggered by sending the object a single \textbf{bang} message or by sending a value to 
their first, or ``hot", inlet. In addition, they may respond to other messages to change internal state
data or configuration - for example the ``set" message is commonly implemented to update internal
state without outputting any result. The end result of object methods run on receipt of a message falls broadly
into three (non-exclusive) categories: the object may update some internal state, it may send a
message or messages out of its outlets, and it may cause a side effect in the broader
Max environment, such as printing to the console or updating a global data element such as
an audio buffer.

@@

Objects are not limited to interacting with other objects through messages passing into inlets and out of 
outlets. A C API exists that enables objects to query and control various engine components (e.g., 
the transport mechanism), and it is also possible for objects to send messages to other objects 
directly or through the scheduler queues, without the sending and receiving objects 
necessarily being connected visually in the patcher. Fundamentally however, the same mechanism
is used - when an object calls a method on another object at the C level, this is still done
by creating a data structure of a message selector and optional atom arguments, 
and then sending this to the receiving object through a generic message sending function.

As each visual box in a patcher has state that is retained between messages, we can see that when
programming the visual patcher (a.k.a. ``patching"), we are in effect programming with
object instantiations rather than classes. In fact, in the Max engine, the act of creating
a new visual-box and placing it on the patcher canvas does indeed instantiate an object, creating
a fresh copy of the prototype's data structure and adding it to a graph of other objects.
This is in contrast to computer music languages such as Csound, where the user programs instruments 
with functions and the engine creates an instantiated data structure on each note-event sent to 
the instrument from a score - in effect the instruments act as object builders and note-events 
become objects that exist for the duration of the note played \cite{Lazzarini2013}.

@@

Scheme For Max is far from the first Lisp-based computer music tool, or even
the first real-time music tool in Lisp. There is a rich history of Lisp in music, 
ranging from Common Lisp Music, created by Bill Schottstaedt in the late 1980's
\cite{Wang17} and Heinrich Taube's Common Music in 1991 \cite{Taube91},
to Dannenberg's Nyquist in 1997 \cite{Dannenberg97},
and Impromptu (later rewritten as Extempore) \cite{Sorensen2010}. 
Nor is Scheme For Max unique in providing a Lisp
based extension facility to a commercial music platform - the author unknowingly
programmed in Lisp in the early 1990's while using the Cakewalk sequencer,
which came with an embedded Lisp interpreter in the form of the 
Cakewalk Application Language (CAL).

@@

In contrast with S4M, systems such as Common Lisp Music, Common Music, Nyquist, and Extempore
all run as stand-alone applications, with events originating from them either
rendered as audio internally or destined for other systems for rendering to audio. 
While these systems are, like Max, also capable
of receiving gestural input, they must, in essence, ``be their own boss" - they provide
their own scheduler and timing clocks and act as the principal engine from which temporal
events originate. If they are to be used alongside another temporal engine, the two
must be synchronized and events sent back and forth over network connections of some kind.

@@

In contrast to these, S4M embeds a light-weight Scheme interpreter directly in a Max
C external with no intermediate or external runtime necessary. This is made possible by the suitability
of Scheme for the creation of small, portable, and self-contained interpreters.
The s7 interpreter is implemented entirely in ANSI C, is statically linked during compilation of S4M, 
and operates entirely in whatever thread it is run in by the C host.
This means multiple instantiations of the interpreter are possible, and
each \textbf{s4m} object has direct C-level access to the Max C API 
through foreign function interface calls, with the results of said calls being exactly 
the same as if the functions were called in C, save execution time. 
The ramification of this is that S4M is unique among Max Lisp projects in being
able to operate within the native Max scheduling system, enabling highly accurate
timing and retaining deterministic control flow within a patcher. (i.e., There
is no hidden networking layer with communication to an outside process, which 
makes operations fundamentally asynchronous even if this is hidden from the user visually.)
This provides the user with opportunities to use Scheme for tasks beyond
what is possible with pre-existing options. For example, we are in effect 
able to say ``at a certain time, read some shared state and update an 
internal engine parameter accordingly". In the example below, a Scheme function
is scheduled that will, in 4 bars time, read a rehearsal mark from the table \textbf{mark-table},
advancing the global transport to that position. As the table
is a shared data object, other Max objects (or patchers) could write to this
rehearsal mark table without awareness of S4M.

@@

@@

The s7 interpreter is implemented in a single C file (with header file), 
and is statically linked with the C code of the \textbf{s4m} object. 
The s7 foreign function interface enables one to define new Scheme functions in C 
for calling into C from Scheme, and to call any Scheme code from C. 
The interpreter is instantiated
in the \textbf{s4m} object's constructor (\textbf{s4m\_new} as per the Max naming conventions), and
a reference to it is stored in the \textbf{s4m} object's data structure.
Additionally, during initialization the external creates a Scheme-side variable 
that holds a pointer to the instantiated \textbf{s4m} object itself. Thus
both C functions that are called from Scheme and C functions called in response to Max
messages have access to both the \textbf{s4m} Max object and its associated s7 interpreter.
There is always one and only one s7 interpreter per \textbf{s4m} object instance, and
as s7 is thread safe, many \textbf{s4m} objects can be created in a Max patch, including
in both the low and high priority threads. (Max runs both a high-priority ``scheduler" 
thread and a low-priority ``UI" thread for event/message flow. Each \textbf{s4m} object 
runs in only one of these, chosen via an optional constructor argument, and incoming messages
from the opposite thread are deferred or promoted accordingly.)

@@

Music is fundamentally a temporal art form - there may be music with static pitch
or static amplitude, but absent rhythm, there is no music.  
Thus a core problem in computer music programming environments is that of providing a flexible 
means of triggering events in time. Further, in platforms intended to support both
live and algorithmic musical interaction, a viable solution must enable the performer
to interact with scheduled events easily, both programmatically and gesturally.

@@

A larger problem is that of how to capture gestural values for use in the delayed
events. Max's visual patching language is fundamentally modelled similarly to a modular
synthesizer - there is one instance in memory of each visual object, and messages can only
pass through them one at a time. Creating visual programs where execution of 
delayed events will trigger cascades of messages through objects that are also 
being used in the interim becomes notoriously complex and there exists no straightforward
facility to express \textit{``make a new container for this variable as it is now so that
we can fetch it later"}. A naive solution would use a visual patcher object to store 
the variable's state at trigger time for use later, however this only works
if there is only one scheduled event - adding more scheduled events requires adding
an additional storage objects per concurrently scheduled event, and as each object in the visual patcher
represents an object instantiation rather than a class, this rapidly becomes
baroque. This ``programming-with-instances" paradigm
is very convenient when we want to ensure there is only ever one instance of a given function, such
as in emulation of a hardware synthesizer, where each oscillator represents one ``always-on"
physical device that is playing regardless of whether one can hear it. 
However, it is difficult in a polyphonic time context
such as the creation and eventual playback of some arbitrary number of delayed events.

@@

In addition to output latency, there is the factor of internal latency.
Max, like many computer music programming environments, generates audio 
in vectors of samples, with
each vector generated from one processing pass, and (normally) control
rate passes conducted also a maximum of once per audio vector pass \cite{Puckette2002}.
The size of this signal vector thus has a large effect on performance, with larger
vectors requiring significantly less CPU use. As audio rendering of note onsets originating
from control messages (the normal scenario) must thus be aligned with 
signal vector boundaries, this produces delays of note onsets by up to
the time period of one vector, a phenomenon known as ``timing jitter" in the nomenclature.
However, the more recent scheduling facilities in Max, including the 
clock facilities used by S4M, are implemented to compensate for this jitter,
maintaining long-term temporal accuracy despite note onset delays. The mechanism
by which this is done is unfortunately not publicly documented or available
in open-source code, but one can assume that it is some variant of the common pattern 
whereby scheduled events are aware of when they are supposed to be running
(``logical-time"), and if they in turn schedule subsequent events, the time for these subsequent
events will be shifted to compensate for the jitter, based 
on the logical time of the scheduling event, rather than its
actual (post-jitter) real time \cite{Anderson1986}.
The author also conducted personal tests of 
this compensation, comparing both high and low signal vector sizes to produce
variable delays in note onsets, and it was determined
that so long as the overall output latency is again large enough to allow
the GC to run in time, self-scheduling events in a long piece do stay
accurate in that jitter does not accumulate over the piece - 
note onsets stay accurate to within the delay of one signal vector of samples,
regardless of the length of the piece of music.

@@

Future plans for the project include a version for the open-source
Pure Data platform, similar to Max in many ways and created also by Miller Puckette,
the original author of Max. \textbf{Scheme For Pure Data} is now in beta release, available
as source code, with binary releases planned for July 2021. In addition
to this, experimental work on running Scheme for control rate DSP is planned,
as well as the creation of portability layers in Scheme to allow one to run 
the same Scheme code on S4M, S4Pd, or Common Music (which also uses s7).
There have been successful ports of s7, Pure Data, and Csound to Web Assembly, 
thus it seems likely that work from the project will be useable for web audio as well.
It is hoped that through these additional run-times, the project can become 
a valuable addition to the larger computer music community, enabling
use in contexts where the commercial Max platform is not practical
or financially accessible.

>>>>

In both the fields of live-coding and algorithmic music, the ability for
the performer and composer to schedule events in the future with high temporal
accuracy is of major benefit. A common scenario is that a solo performer or duo
is composing electronic music on the fly for a large number of voices --- more than
the performers have hands. Thus they create abstractions of musical content,
and request from the system that potentially many of these will begin together
at some point in the future. For example, a performer may program a drum
groove that has a large number of interacting lines, but wish the groove
to only begin playing, and thus be audible, at the beginning of the next 16-bar 
musical boundary. While this is possible in the Max visual
patching language, it is cumbersome and thus the implementation is not
optimal for live-coding or algorithmic music where the speed with which
the user can accomplish this is an important consideration. In a live
electronic music context, one of the primary challenges for the perfomers
is the design of a working system that enables them to create new material on
stage (making the performances engaging and improvisatory), 
but in a way that enables them to work fast enough not to bore an audience!

@@

Activity in an object (represented visually by a single visual box in the patch) is triggered
by sending the object a Max message, most commonly from an object connected to it
through a patch cord running to an ``inlet" of the object. 
In order for the receiving object to do anything, it must 
be sent a message with a message selector for which it has a bound method,
a situation referred to in the nomenclature as ``responding to the message". Most objects
have a principal activity that typically ends with outputting the result of their calculation, and this is
often triggered by sending the object a single \textbf{bang} message or by sending a value to 
their first, or ``hot", inlet. In addition, they may respond to other messages to change internal state
data or configuration --- for example the ``set" message is commonly implemented to update internal
state without outputting any result. The end result of object methods run on receipt of a message falls broadly
into three (non-exclusive) categories: the object may update some internal state, it may send a
message or messages out of its outlets, and it may cause a side effect in the broader
Max environment, such as printing to the console or updating a global data element such as
an audio buffer.

@@

Objects are not limited to interacting with other objects through messages passing into inlets and out of 
outlets. A C API exists that enables objects to query and control various engine components (e.g., 
the transport mechanism), and it is also possible for objects to send messages to other objects 
directly or through the scheduler queues, without the sending and receiving objects 
necessarily being connected visually in the patcher. Fundamentally however, the same mechanism
is used --- when an object calls a method on another object at the C level, this is still done
by creating a data structure of a message selector and optional atom arguments, 
and then sending this to the receiving object through a generic message sending function.

As each visual box in a patcher has state that is retained between messages, we can see that when
programming the visual patcher (a.k.a. ``patching"), we are in effect programming with
object instantiations rather than classes. In fact, in the Max engine, the act of creating
a new visual-box and placing it on the patcher canvas does indeed instantiate an object, creating
a fresh copy of the prototype's data structure and adding it to a graph of other objects.
This is in contrast to computer music languages such as Csound, where the user programs instruments 
with functions and the engine creates an instantiated data structure on each note-event sent to 
the instrument from a score --- in effect the instruments act as object builders and note-events 
become objects that exist for the duration of the note played \cite{Lazzarini2013}.

@@

Scheme For Max is far from the first Lisp-based computer music tool, or even
the first real-time music tool in Lisp. There is a rich history of Lisp in music, 
ranging from Common Lisp Music, created by Bill Schottstaedt in the late 1980's
\cite{Wang17} and Heinrich Taube's Common Music in 1991 \cite{Taube91},
to Dannenberg's Nyquist in 1997 \cite{Dannenberg97},
and Impromptu (later rewritten as Extempore) \cite{Sorensen2010}. 
Nor is Scheme For Max unique in providing a Lisp-family
extension facility to a larger music platform --- the author unknowingly
programmed in Lisp in the early 1990's while using the Cakewalk sequencer,
which came with an embedded Lisp interpreter in the form of the 
Cakewalk Application Language (CAL)! There also exist Lisp-family client interfaces
to the open-source SuperCollider music platform, including Overtone (Clojure),
rsc3 (r6rs Scheme), and cl-collider (Common Lisp).

@@

In contrast with S4M, systems such as Common Lisp Music, Common Music, Nyquist, and Extempore
all run as stand-alone applications, with events originating from them either
rendered as audio internally or destined for other systems for rendering to audio. 
While these systems are, like Max, also capable
of receiving gestural input, they must, in essence, ``be their own boss" --- they provide
their own scheduler and timing clocks and act as the principal engine from which temporal
events originate. If they are to be used alongside another temporal engine, the two
must be synchronized and events sent back and forth over network connections of some kind.

@@

In contrast to these, S4M embeds a light-weight Scheme interpreter directly in a Max
C external with no intermediate or external runtime necessary. This is made possible by the suitability
of Scheme for the creation of small, portable, and self-contained interpreters.
The s7 interpreter is implemented entirely in ANSI C, is statically linked during compilation of S4M, 
and operates entirely in whatever thread it is run in by the C host.
This means multiple instantiations of the interpreter are possible, and
each \textbf{s4m} object has direct C-level access to the Max C API 
through foreign function interface calls, with the results of said calls being exactly 
the same as if the functions were called in C, save execution time. 
The ramification of this is that S4M is unique among Max Lisp projects in being
able to operate within the native Max scheduling system, enabling highly accurate
timing and retaining deterministic control flow within a patcher. (i.e., There
is no hidden networking layer with communication to an outside process, which 
makes operations fundamentally asynchronous even if this is hidden from the user visually.)
This provides the user with opportunities to use Scheme for tasks beyond
what is possible with pre-existing options. For example, we are in effect 
able to say ``at a certain time, read some shared state and update an 
internal engine parameter accordingly". In the example below, a Scheme function
is scheduled that will, in 4 bars time, read a rehearsal mark from the table \textbf{mark-table},
advancing the global transport to that position. As the table
is a shared data object, other Max objects (or patchers) could write to this
rehearsal mark table without awareness of S4M. (Note: this is a safe operation
from Scheme code as table access is done through thread-safe Max SDK functions 
that handle synchronization and locking for us.)

@@

\section{rationale for the choice of s7 scheme}

The design of the Max external system means that, technically, any interpreter
that can be embedded in a C host could be used. The author evaluated several during
the initial prototyping stages of development, including ECL (Embeddable Common Lisp),
Gambit, Chicken, Racket, Guile, and s7. The choice to use s7 was based on several 
factors, including the ease of embedding, its license, and its use in other computer music tools (and
thus its design suitability for computer music).

The s7 interpreter is implemented in a single C file (with header file), 
and is statically linked with the C code of the \textbf{s4m} object. Embedding is
as simple as linking to the s7 library and including a single file. In this 
regard, it is similar to Guile, and in fact, is linguistically similar to Guile 1.8.
However, while Guile is reciprocally licensed with the GPL, s7 is liberally licensed 
under the BSD license. As the creation and sale of commercially externals is part of the Max 
milieu, the author felt this was more likely to result in uptake within the Max community.

Racket, Chicken, and Gambit were also potential contenders. In all three cases, the
complexity of embedding was significantly higher, and in at least one case (Chicken),
thread-safety for multiple interpreters in the same Max patch appeared problematic.
In addition, documented examples of common embedding use cases were less available
than with s7.

Finally, s7 was created for computer music, and as a result is used in other computer music
environments, such as Common Music. As a result, there is a rich body of code, especially
from Common Music, that can be used or studied by users. There is also a small, but active, 
online community of music programmers using s7, communicating through the CM-dist mailing
list hosted at the Stanford computer music center, CCRMA. The author of s7, Bill
Schottstaedt, has been an enthusiastic supporter of the project and has been invaluable
with technical assistance.

@@

The s7 foreign function interface enables one to define new Scheme functions in C 
for calling into C from Scheme, and to call any Scheme code from C. 
The interpreter is instantiated
in the \textbf{s4m} object's constructor (\textbf{s4m\_new} as per the Max naming conventions), and
a reference to it is stored in the \textbf{s4m} object's data structure.
Additionally, during initialization the external creates a Scheme-side variable 
that holds a pointer to the instantiated \textbf{s4m} object itself. Thus
both C functions that are called from Scheme and C functions called in response to Max
messages have access to both the \textbf{s4m} Max object and its associated s7 interpreter.
There is always one and only one s7 interpreter per \textbf{s4m} object instance, and
as s7 is thread safe, many \textbf{s4m} objects can be created in a Max patch, including
in both the low and high priority threads. (Max runs both a high-priority ``scheduler" 
thread and a low-priority ``UI" thread for event/message flow. Each \textbf{s4m} object 
runs in only one of these, chosen via an optional constructor argument, and incoming messages
from the opposite thread are deferred or promoted accordingly.)

@@

Music is fundamentally a temporal art form --- there may be music with static pitch
or static amplitude, but absent rhythm, there is no music.  
Thus a core problem in computer music programming environments is that of providing a flexible 
means of triggering events in time. Further, in platforms intended to support both
live and algorithmic musical interaction, a viable solution must enable the performer
to interact with scheduled events easily, both programmatically and gesturally.

@@

A larger problem is that of how to capture gestural values for use in the delayed
events. Max's visual patching language is fundamentally modelled similarly to a modular
synthesizer --- there is one instance in memory of each visual object, and messages can only
pass through them one at a time. Creating visual programs where execution of 
delayed events will trigger cascades of messages through objects that are also 
being used in the interim becomes notoriously complex and there exists no straightforward
facility to express \textit{``make a new container for this variable as it is now so that
we can fetch it later"}. A naive solution would use a visual patcher object to store 
the variable's state at trigger time for use later, however this only works
if there is only one scheduled event --- adding more scheduled events requires adding
an additional storage objects per concurrently scheduled event, and as each object in the visual patcher
represents an object instantiation rather than a class, this rapidly becomes
baroque. This ``programming-with-instances" paradigm
is very convenient when we want to ensure there is only ever one instance of a given function, such
as in emulation of a hardware synthesizer, where each oscillator represents one ``always-on"
physical device that is playing regardless of whether one can hear it. 
However, it is difficult in a polyphonic time context
such as the creation and eventual playback of some arbitrary number of delayed events.

@@

In addition to output latency, there is the factor of internal latency.
Max, like many computer music programming environments, generates audio 
in vectors of samples, with
each vector generated from one processing pass, and (normally) control
rate passes conducted also a maximum of once per audio vector pass \cite{Puckette2002}.
The size of this signal vector thus has a large effect on performance, with larger
vectors requiring significantly less CPU use. As audio rendering of note onsets originating
from control messages (the normal scenario) must thus be aligned with 
signal vector boundaries, this produces delays of note onsets by up to
the time period of one vector, a phenomenon known as ``timing jitter" in the nomenclature.
However, the more recent scheduling facilities in Max, including the 
clock facilities used by S4M, are implemented to compensate for this jitter,
maintaining long-term temporal accuracy despite note onset delays. The mechanism
by which this is done is unfortunately not publicly documented or available
in open-source code, but one can assume that it is some variant of the common pattern 
whereby scheduled events are aware of when they are supposed to be running
(``logical-time"), and if they in turn schedule subsequent events, the time for these subsequent
events will be shifted to compensate for the jitter, based 
on the logical time of the scheduling event, rather than its
actual (post-jitter) real time \cite{Anderson1986}.
The author also conducted personal tests of 
this compensation, comparing both high and low signal vector sizes to produce
variable delays in note onsets, and it was determined
that so long as the overall output latency is again large enough to allow
the GC to run in time, self-scheduling events in a long piece do stay
accurate in that jitter does not accumulate over the piece --- 
note onsets stay accurate to within the delay of one signal vector of samples,
regardless of the length of the piece of music.

@@

Future plans for the project include a version for the open-source
Pure Data platform, similar to Max in many ways and created also by Miller Puckette,
the original author of Max. \textbf{Scheme For Pure Data} is now in beta release, available
as source code, with binary releases planned for August 2021. In addition
to this, experimental work on running Scheme for control-rate DSP is planned.
This would entail Scheme executing in the timed audio rendering thread, but
at a subdivision of the audio sample rate such as every 32 or 64 samples, 
thus generating streams similar to the way digital audio is rendered, but
at a lower resolution suitable for use as signal modulators. (The ability
to run DSP code at a lower sample rate for modulator signals is  
is called the "control rate" in CSound and other similar audio programming platforms.)

The author also plans to create portability layers in Scheme to allow one to run 
the same Scheme code on S4M, S4Pd, or Common Music (which also uses s7),
and potentially in C++ projects through the JUCE audio development framework.
There have also been successful ports of s7, Pure Data, and Csound to Web Assembly, 
thus it seems likely that work from the project will be useable for web audio as well.
It is hoped that through these additional run-times, the project can become 
a valuable addition to the larger computer music community, enabling
use in contexts where the commercial Max platform is not practical
or financially accessible.

As s7 Scheme is open-source itself, another potential area of work
is improving s7's suitability for real-time use. This could include
implementation of an incremental garbage collector, for example.
